---
title: Looking for non english package in CRAN
author: Christophe Dervieux
date: '2018-03-11'
description: An experiment of detecting language in DESCRIPTION
slug: non-english-pkg-in-cran
categories:
  - R
tags:
  - R
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Trying to figure out how encoding work in `r emo::ji("package")` for writing french DESCRIPTION and manual files, I was wondering if others packages on CRAN as been wrote in French or any other non english language.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Looking into the cran database

First, we load the cran database with all information about description fields. We keep only some fields that can helps us look for not english packages. (For a list of all the fields meaning, read [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-DESCRIPTION-file))

```{r}
to_keep  <- c("Package", "Title", "Description", "Encoding", "Language")
cran_db <- tools::CRAN_package_db()[to_keep] %>% as_tibble()
glimpse(cran_db)
```

Let's see how many package set their Encoding field in their DESCRIPTION

```{r}
nb_of_na <- cran_db %>% filter(is.na(Encoding)) %>% count(Encoding) %>% pull(n)
ggplot(cran_db %>% filter(!is.na(Encoding))) +
  geom_bar(aes(x = Encoding), fill = "skyblue") +
  labs(title = "Nb of package with Encoding field", subtitle = glue::glue("{nb_of_na} packages with unset Encoding field")) +
  theme_light()
```

Is there any package that uses a Language field in their DESCRIPTION ? 

```{r}
nb_of_na <- cran_db %>% filter(is.na(Language)) %>% count(Language) %>% pull(n)
ggplot(cran_db %>% filter(!is.na(Language))) +
  geom_bar(aes(x = Language), fill = "skyblue") +
  labs(title = "Nb of package with Language field", subtitle = glue::glue("{nb_of_na} packages with unset Language field")) +
  coord_flip() +
  theme_light()
```

So there are some package that declares their Language field in the DESCRIPTION. These field means: 

> A ‘Language’ field can be used to indicate if the package documentation is not in English: this should be a comma-separated list of standard (not private use or grandfathered) IETF language tags as currently defined by RFC 5646 (https://tools.ietf.org/html/rfc5646, see also https://en.wikipedia.org/wiki/IETF_language_tag), i.e., use language subtags which in essence are 2-letter ISO 639-1 (https://en.wikipedia.org/wiki/ISO_639-1) or 3-letter ISO 639-3 (https://en.wikipedia.org/wiki/ISO_639-3) language codes. 

Which are those packages ? 
```{r, results='asis'}
cran_db %>%
  filter(Language == "fr") %>%
  pander::pandoc.table(split.tables = Inf)
```

We see that neither of these packages have their title or description field in
french. Moreover, one package has UTF8 encoding and the other Latin1.

So this Language field is not very useful. I am not sure if this field is useful
for any other purpose than information

# Trying to detect languages of Packages

How can we have some insight on CRAN package language ? 

We can analyse the cran database using the Google's Compact Language Detector. ROpensci has made 2 wrappers for using it in version 2 ([`cld2`](https://github.com/ropensci/cld2)) and version 3 ([`cld3`](https://github.com/ropensci/cld3)). 

We will be using the version 2, as `cld3` is still experimental. 

```{r}
cran_db_with_lang <- cran_db %>%
  mutate(detected = cld2::detect_language(Description, lang_code = FALSE))
```


```{r}
nb_of_lg_en <- cran_db_with_lang %>% filter(detected == "ENGLISH") %>% count(detected) %>% pull(n)
ggplot(cran_db_with_lang %>% filter(detected != "ENGLISH")) +
  geom_bar(aes(x = detected), fill = "skyblue") +
  labs(title = "Language of packages", subtitle = glue::glue("{nb_of_lg_en} packages detected as English")) +
  coord_flip() +
  theme_light()
```

There are very few packages that seems to have non english description field.
```{r, echo = FALSE, results='asis'}
cran_db_with_lang %>% 
  filter(detected != "ENGLISH") %>%
  pander::pandoc.table(split.tables = Inf)
```

The detection by `cld2` does not seem to be 100% correct. 

However, it helps us isolate some non english packages and it highlights that
English is the language to go to avoid encoding issues and specific
configurations in package. 
I will now continue to look into those packages I just found to see if they have
the same encoding issue as French package... and it seems so !
